\copyrightyear{2026}
\copyrightclause{Copyright for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0).}

\conference{ITAT'26: Information technologies -- Applications and Theory, September 25--29, 2026, Vršatské Podhradie, Slovakia}

\title{Towards Transfer Learning for Hyperparameter Optimization}

\author[1]{Marek Dědič}[%
orcid=0000-0003-1021-8428,
email=marek@dedic.eu,
url=https://dedic.eu/,
]
\cormark[1]
\address[1]{Czech Technical University in Prague, Trojanova 13, Prague, Czech Republic}

\author[2]{Martin Holeňa}[%
email=martin@cs.cas.cz,
url=https://www.cs.cas.cz/~martin/,
]
\address[2]{Czech Academy of Sciences, Pod Vodárenskou věží 271/2, Prague, Czech Republic}

\cortext[1]{Corresponding author.}

\begin{abstract}
  Hyperparameter optimization (HPO) for Graph Neural Networks (GNNs) remains computationally expensive, and the Cross-RF meta-learning framework has shown promise in accelerating the search by transferring knowledge from prior runs across datasets. However, two key limitations restrict its practical applicability: the requirement for an existing library of completed HPO runs as pre-training data, and its current role as a one-shot warm-starter rather than an active participant in the optimization loop. This paper studies both limitations. First, we investigate pre-training the Cross-RF metamodel on synthetic graph datasets -- generated via Erdős–Rényi and stochastic block models -- as a substitute for real prior runs, enabling effective cold-start transfer learning without any prior experimentation on real graphs. Second, we propose integrating the Cross-RF metamodel as a surrogate within a Sequential Model-Based Optimization (SMBO) framework, allowing it to guide the acquisition function throughout the optimization rather than only at initialization.
  \todo{Update}
\end{abstract}

\begin{keywords}
  hyperparameter optimization \sep
  metalearning \sep
  transfer learning \sep
  graph neural networks
\end{keywords}

\maketitle
